Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-11-14 paper/paper.tex
remove outdated note
lambek calculus related work

The design of \theoryabbv
is an extension of Joachim Lambek's \emph{categorial grammars}
\cite{lambek58}. Categorial grammars are a grammar formalism equivalent
in expressive power to context-free grammars that in modern
terminology would be considered a kind of \emph{non-commutative linear
logic} --- a version of linear logic where the tensor product is not
commutative, reflecting the obvious property that the relative
ordering of characters is significant in parsing problems. We extend
non-commutative linear logic with two key components that increase its
power to support arbitrarily powerful grammar formalisms: inductive
linear types, as well as dependency of linear types on non-linear
data. The resulting system has two kinds of types: non-linear types
which model sets and linear types which model formal
grammars. Crucially, the non-linear types and linear types are allowed
to be dependent on non-linear types, but not on linear types. This
combination has been used previously in the ``linear-non-linear
dependent'' type theory with \emph{commutative} linear logic to model
imperative programming \cite{krishnaswami_integrating_2015}.

@@

\subsubsection{Lambek Calculus and Categorial Grammar}

\steven{Note that lambek calculus and its extensions don't have inductives, or
  even kleene star as far as I can tell. }

\steven{Choppy notes about lambek calc:
In 1958 \cite{lambek58}, Joachim Lambek introduced his syntactic
calculus as a logical system for linguistic derivations. In fact, one may view
the subtheory of our calculus generated over $\otimes, \lto, \tol$ to be a proof
relevant presentation of Lambek's original system.

Lambek later explicitly connects his syntactic calculus to the structure of a
non-commutative biclosed monoidal category \cite{lambek1988categorial}. There is
a rich line of subsequent research after these revelations by Lambek, including
the study of pregroup grammars
\cite{coeckeMathematicalFoundationsCompositional2010}, the presentation of
abstract categorial grammars \cite{degrooteAbstractCategorialParsing2015}, and
even Lambek calculus with dependent types \cite{luoLambekCalculusDependent}. As
far as the authors can tell, these works seem to be primarily interested in natural
language semantics in the style of Montague grammars, rather than parsing formal
languages. That is, they seem to try to give a functorial semantic
interpretation of natural language.

There seem to be some implementations from the abstract categorial grammar
toolkit, but again they seem natural language focused and not up to the task of
verified formal language parsing.

Also mention here that Luo's work \cite{luoLambekCalculusDependent} seems similar at first
glance, but isn't as similar as it might look because the judgments and
dependent types seem to
take on a different interpretation of ours. Say something here like, ``as we
understand it, it differs in x,y,z ways''
}

@@

  \steven{How does the distributivity morphism handled? Seems like its a barrier to invoking coherence}

>>>>

The design of \theoryabbv
is an extension of Joachim Lambek's \emph{syntactic calculus}
\cite{lambek58}. Lambek calculus is a grammar formalism equivalent
in expressive power to context-free grammars that in modern
terminology would be considered a kind of \emph{non-commutative linear
logic} --- a version of linear logic where the tensor product is not
commutative, reflecting the obvious property that the relative
ordering of characters is significant in parsing problems. We extend
non-commutative linear logic with two key components that increase its
power to support arbitrarily powerful grammar formalisms: inductive
linear types, as well as dependency of linear types on non-linear
data. The resulting system has two kinds of types: non-linear types
which model sets and linear types which model formal
grammars. Crucially, the non-linear types and linear types are allowed
to be dependent on non-linear types, but not on linear types. This
combination has been used previously in the ``linear-non-linear
dependent'' type theory with \emph{commutative} linear logic to model
imperative programming \cite{krishnaswami_integrating_2015}.

@@

\subsubsection{Lambek Calculus and Categorial Grammar}
Lambek's original syntactic calculus \cite{lambek58} describes a logical system
for linguistic derivations. Later, he explicitly gives semantics to his
syntactic calculus inside of a non-commutative biclosed monoidal category
\cite{lambek1988categorial}. Following Lambek's illuminating ideas, Buszkowski
gives a succinct overview of the
rich body of work studying the role of linear logic and lambda calculi in
linguistics \cite{buszkowskiTypeLogicsGrammar2003}. \steven{Is citing a survey
  like this okay?} Grammars based on logic in
this manner may go by the names Lambek calculus, typelogical grammar, or
categorial grammar. Such categorial grammars correspond to the subtheory of our
calculus generated by $\otimes$, $\lto$, and $\tol$.

While applicable to both
formal and natural languages, the existing works on categorial grammar are
frequently oriented toward giving functorial semantics to natural language in
the tradition of Montague \cite{Montague1974-MONFPS}. The Abstract Categorial
Grammar Toolkit \cite{Guillaume2024} and Grammatical Framework \cite{ranta-2011}
each mechanize aspects of Lambek calculus ; however, to the best of our knowledge,
\theoryabbv is the first mechanization designed for formal language parsing.

\steven{Contrast their interpretation of dependent types with ours. How tf do I
  say idk what they're doing. Perhaps just say that their dependent types take
  on a different interpreation and no more.}
Luo \cite{luo} describes a Lambek calculus with dependent types.

@@
